Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-02-19 supplementary.tex
reduce pages by restructuring figures and moving proof of lemma to supplimentary

\usepackage{setspace, url}
\doublespacing

@@

% Article top matter
\title{Validation of Visual Statistical Inference, Applied to Linear Models}
\author{
{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\thanks{Mahbubul Majumder is Graduate Student (e-mail: mahbub72@gmail.com) , Heike Hofmann is Associate  Professor and Dianne Cook is Professor, Department of
Statistics and Statistical Laboratory, Iowa State University,
Ames, IA 50011-1210. This research is supported in part by the
National Science Foundation Grant \# DMS 1007697.}}
\date{\vspace{-.7in}}
%\date{\today}  %\today is replaced with the current date
\maketitle

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently, there were no formal visual methods in place for determining statistical significance of findings. This changed when \citet{buja:2009} conceptually introduced the lineup protocol for formal tests of visual findings. In this paper this is taken a step further by refining the terminology of visual inference, framing the protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists, and it is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.

{\bf Keywords: \sf statistical graphics, lineup, non-parametric test, data mining, visualization, exploratory data analysis} 
\end {abstract}

%\begin{multicols}{2}
%\twocolumn

\section{Introduction}

%\checkmark

Statistical graphics nourish the discovery process in data analysis by revealing unexpected things,  finding structure that was not previously anticipated,  or orthogonally by contrasting prevailing hypotheses. The area of graphics is often associated with exploratory data analysis, which was pioneered by \cite{tukey:eda} and, is particularly pertinent in today's data-rich world where discovery during data mining has become an important activity. Graphics are also used in many places where numerical summaries simply do not suffice: model checking, diagnosis, and in the communication of findings.

Statistical graphics research and development is agreeably established now, and has been cemented by several new developments in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. This body of work provides a framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allows the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between conventional statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the actual data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? (The position of the actual data plot is provided in Section \ref{sec:category}.) Wrapped in a process that mirrors conventional inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the actual data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done based on a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

%, analysis requires plots of data. For exploratory data analysis, statistical graphics play an invaluable role in model checking and diagnostics. Even though we have established mathematical procedures to obtain various statistics, we need to support the results by also producing the relevant plots.

%The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS ease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}.   \blue {Graham Wills has implemented the grammar in SPSS, but I'm not sure how to cite this.}

%\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between conventional statistical inference procedures and exploratory data analysis.

%In this paper we compare the lineup protocol to conventional statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. The effect of the lineup size and number of observers on the power of the test is discussed in Section \ref{sec:size}. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes  three  user studies based on simulation experiments conducted to compare the power of the lineup protocol with the equivalent conventional test and  Section \ref{sec:results} presents an analysis of the resulting data. %The last section describes the next steps in this research.

\section{Definitions and Explanations for Visual Statistical Inference} \label{sec:visual_test}

Table \ref{tbl:compare} illustrates the lineup protocol in relation to conventional hypothesis testing. Both start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

\begin{table*}[hbtp]
\caption{Comparison of visual inference with conventional inference.}
\centering 
\begin{tabular}{llll} 
\hline \hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

 Test statistic & $T(y)=\frac{\hat{\beta}}{se(\hat{\beta})}$ & $T(y)=$ \begin{minipage}[h]{1cm} \begin{center} \scalebox{0.45}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} \\

 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

 Sampling Distribution & $f_{T(y)}(t); $\begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.55}{\includegraphics{stat_mathematical_test.pdf}} \end{center} \end{minipage} & $f_{T(y)}(t); $ \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.32}{\includegraphics{lineup_category_small.pdf}} \end{center} \end{minipage} \\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\
 Reject $H_0$ if & actual $T$ is extreme & actual plot is identifiable \\
\hline 
\end{tabular}
\label{tbl:compare}
\end{table*}

%Visual tests are non-parametric tests of the form $H_0$: the data plot is visually not distinguishable from a null plot vs $H_1$: the data plot is visually distinguishable from a plot generated consistently with $H_0$.

%Power of a plot and $p$-value have been derived in Buja and Wickham as ...

%for this paper, we want to place the parametric inference of a regression setting into the framework of visual testing:

%This section outlines the concepts of visual inference in comparison to the procedures of conventional statistical inference. Table \ref{tbl:compare} (derived from \citet{buja:2009}) gives a summarized overview of this comparison.

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that maps the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T(y_0)$ which would be possible values of $T$ if $\theta \in \Theta$.  %Similarly, for any test statistic, we have a rejection region $R$ and, complementary to that, $R^c$,  that partitions the space of the test statistic, which is usually a subspace of the real numbers. 
For visual inference, unlike in the conventional hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the actual data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

\begin{dfn} \label{dfn:test}
A \textbf{visual test statistic}, $T(.)$, is a function of a sample that produces a plot. $T(y)$ maps the actual data to the plot, and we call this the \textbf{(actual) data plot}, and $T(y_0)$ maps a sample drawn from the null distribution into the same plot form. These type of plots are called \textbf{null plots}. 
\end{dfn}

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. \\
\noindent
The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

\begin{dfn}\label{dfn:lplot}
A \textbf{lineup} is a layout of $m$ randomly placed visual statistics, consisting of 
\begin{itemize}\itemsep-3pt
\item $m-1$ statistics, $T(y_0)$, simulated from the model specified by $H_0$  (null plots) and 
\item the test statistic, $T(y)$, produced by plotting the actual data, possibly arising from $H_1$.
\end{itemize}
\end{dfn}

\noindent The $(m-1)$ null plots are % can be 
members of the sampling distribution of the test statistic assuming that the null hypothesis is true. If $H_1$ is true, we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots. A careful visual inspection of the lineup by %\strikethrough{an} 
independent observers follows;  observers are asked to point out the plot most different from the lineup. If the test statistic is identified in the lineup, this is considered as evidence against the null hypothesis. 
%If it is not identified the conclusion is to not reject the null hypothesis.

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}_{K, 1/m}$,  therefore:

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ evaluated by $K$ observers is  given as

\begin{equation}\label{binom}
P(X \ge x) = 1 - \text{Binom}_{{K, 1/m}} (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i}
\end{equation}

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\scalebox{.9}{
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\ \hline %\cline{1-3}\cline{4-6}\cline{7-9}\cline{10-12}\cline{17-19}
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &           &   &        && 4 & 4 & $< 0.0001$ & 5 & 4 & $< 0.0001$ \\\hline
\end{tabular}
}
\end{center}
\label{pvalue}
\end{table}

%\begin{dfn}\label{dfn:test}
%\red{why do we need the $\theta$? - it's not defined and not explained. usually we would assume that lineups are non-parametric}
%The \textbf{visual test}, $V_{\theta}$, is defined to be 
%\begin{itemize}\itemsep-3pt
%\item \textbf{Reject} $H_0$ if `enough' observers select the actual data plot in the lineup of $m$ plots, and
%\item \textbf{Fail to reject} $H_0$  otherwise. %if the observer selects a null plot.
%\end{itemize}
%where `enough' observers is defined as some threshold $x_{1-\alpha}$, for which $P(X \ge x_{1-\alpha}) < \alpha$, the level of significance.
%\end{dfn}

%\blue{To avoid confusion with one tailed $(1-\alpha)$ vs two tailed $(1-\alpha/2)$ notation, can we just use $x_{\alpha}$? }

%\red{alternative definition of visual test where $type-I$ error $\alpha$ is attached.}

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of  size $m$ and significance level $\alpha$, is defined to  
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot in the lineup, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
\end{dfn}

\noindent
Associated with any test is the risk of  Type I or II errors, which for visual inference are defined as follows:

\begin{dfn}\label{dfn:error}
The \textbf{Type I error} associated with visual test $V_{\theta}$ is the probability of rejecting $H_0$ when it is true; the probability for that is $P(X \ge x_{\alpha})$, which is controlled by $\alpha$.\\ \noindent
The \textbf{Type II error} is the probability of failing to identify the actual data plot, when $H_0$ is not true, $P( X <  x_{\alpha})$.
\end{dfn}

%\green{Oh, we are using observed and observer, which might be confusing. May need to change observed data plot to something else. Will think about this. Will use ``actual'' for now.} \hh{I think that change is done}

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$. %, with an exact $x_{\alpha}$. 
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p_i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition (see \citet{butler93} for details). Ways to estimate $p_i$ will be  discussed in the following sections.
%one of the main concerns for the remainder of the paper. One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

%\green{WHY??? And how does this relate to the logistic model below?} 
%This leads to the definition of the power of the visual test, as follows:

%The lineup plot can be evaluated by one or more individuals. When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

 %Notice that when $N=1$, this $p$-value is $\frac1m$.

%\subsection{Power}

%Next, we will develop power theoretically and then relate it to the empirical results.
%We will first theoretically define and  derive the power of  a visual test in the setting of a parametric hypothesis test, i.e. we assume a test  of $H_0: \theta = 0$ vs $H_1: \theta \neq 0$ for some real valued parameter vector $\theta$. In a next step we will then relate the theoretical results to our empirical findings.

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test, $V_{\theta}$, is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0 \mid \theta) 
%      = P(X \ge x_{\alpha}) =
%        \begin{cases} 
%               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true}.\\
%              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
%        \end{cases}
    \end{equation*}
\end{dfn}

\noindent An important difference between conventional and visual testing is that lineups will depend on observers evaluation. Thus, the number of observers affects the estimation of power, where $X$ is the number of observers (out of $K$) who identify the actual data plot from the lineup, and power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
Here $F_{X, \theta}$ is the distribution of $X$ and $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. Note that the distribution  $F_X$ depends on which hypothesis is true:
under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
Power_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
%where $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
If the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's skill to identify the actual data plot will require that an individual evaluates multiple lineups.

%\noindent Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.

%\red{there may be a flow in this definition. Due to the discrete nature of binomial (specially $x_{\alpha}$) this definition would only give a conservative lower limit of the actual power. I feel like we should not define power using binomial.}

Power is an important consideration in deciding which test to use for solving a problem. %In conventional testing, new test procedures are examined for their power under a variety of scenarios. 
Here we use it to compare the performance of the visual test with the conventional test, but in practice for visual inference it will mostly be important in deciding on the choice of plot to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. %Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

\section{Effect of Observer Skills and Lineup Size}~\label{sec:size}

%\green{Notation changes $i$ to $j$, and $p$ to $\pi$. Can someone please decide which to use, and make the appropriate changes? How does the ``Let $X_{j\ell} \sim B_{1, \pi_{j\ell}}$...'' below relate to the ``let $X_i$ be a binary random variable...'' above? It looks like this is a repeat. One needs to go once notation is clarified.}

%\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate actual data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

\subsection{Subject-specific abilities}~\label{sec:model}
%Calculating the power  is intricately related to the probability to identify the actual data plot from the lineup.

Suppose each of $K$ independent observers gives evaluations on multiple lineups, and responses are considered to be binary random variable, $X_{\ell i} \sim  Binom_{1, p_{\ell i}}$, where $X_{\ell i} = 1$, if subject $i$ correctly identifies the actual data plot on lineup $\ell$, $1 \le \ell \le L$, and 0 otherwise. A mixed effects logistic regression model is used for $P(X_{\ell i} = 1) =  p_{\ell i} = E(X_{\ell i})$, accommodating both for different abilities of observers as well as differences in the difficulty of lineups.

%Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the actual data plot from lineup $\ell$, and 0 otherwise.
The model can be  fit as:% $p_{ \ell i}$ with a logistic regression with subject-specific random effects as:
\begin{equation} \label{eqn:mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

\noindent where $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well lineup-specific elements, e.g. effect size or difficulty level.
%
$Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups. 
Note that $\tau_{\ell i }$ usually only includes a partial interaction; for a full interaction of subjects' skills and lineup-specific difficulty we would need replicates of the same subject evaluating the same lineup, which in practice is not feasible without loosing independence.

\noindent Using the inverse {\it logit} link function, $g^{-1}(.)$, from  Equation \ref{eqn:mixed} leads to the estimate of the subject and the lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} =  g^{-1}(W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}).
\end{equation}

%\red{This model is  described in more detail in equation \eqref{eqn:mixed}.}

%\noindent For any single observer, the power of a visual test has a lower limit of $\frac1m$ since the probability that an observer will randomly pick the test statistic under $H_0$ is $\frac1m$. When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability 
%\begin{equation}\label{eqn:pvalue}
%Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}
%\end{equation}
%where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $V_{\theta}$ as follows:
%\begin{dfn} \label{dfn:pval}
%The $p$-value associated with the decision of a visual test $V_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
%where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$ as given in eqn (\ref{eqn:pvalue}). Since $u$ is a discrete value with boundaries 0 and $K$,
% we can also re-formulate the $p$-value in terms of upper and lower boundaries:
% \begin{equation*}
% \begin{array}{rcll}
% P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=0\\
%( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
% \end{array}
% \end{equation*}
% }

%\red{HH: Tried to re-write it to answer the question below.
%Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }
%
%    \begin{equation*}
%        \begin{cases} 
%              p\text{-value} \le Pr(U \ge u)    & \text{when reject $H_0$,} \\
%              p\text{-value} > 1-Pr(U \ge u)  & \text{when fail to reject $H_0$}
%        \end{cases}
%    \end{equation*}
%\end{dfn}

%\red{HH: the following paragraph needs some adjustment in terms of t and $\theta$ - could you try to take of that, Mahbub?}\\
%\red{For the following paragraph, we wanted to get rid of the absolute value of $t$ and instead look at $t \in R$ and $t \in R^c$.}

%\noindent
\subsection{Lineup size, $m$}

The finite number $m-1$ of representatives of the null distribution used  as comparison against the test statistic, is a major difference between visual inference and the conventional testing. The choice of $m$ has an obvious impact on the test.

The following properties can only be derived for the situation of a fully parameterized  simulation study, as conducted in this paper. They allow for a direct comparison of lineup tests against the conventional counterparts, and also to identify properties relevant for a quality assessment of lineups when they are used in practical settings. Two assumptions are critical:
\begin{enumerate} \itemsep 0in
\item  the plot setup is structured in a way that makes it possible for an observer to identify a deviation from the null hypothesis,
\item an observer is able to identify the plot with the strongest `signal' (or deviation from $H_0$)  from a lineup.
\end{enumerate}
Evidence in support of the second assumption will be seen in the data from the study discussed in Section~\ref{sec:simulation}, the degree to which the first assumption is fulfilled is reflected by the power of a lineup. The better suited a design is for a particular task, the higher its power will be.

In order to compare the power of conventional and visual tests side-by-side, it is necessary to assume, that we are in the controlled environment of a simulation with tests corresponding to a known parameter value $\theta \in R$ and associated distribution function $F_t$ of the test statistic.

\begin{lemma}~\label{lemma}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the conventional test statistic. Suppose the associated test statistic in a conventional test is observed as  $t_{obs}$ with $p$-value $p_D$.

The probability of picking the data plot from a lineup depends on the size $m$ of the lineup and the strength of the signal in the data plot. 
Under the above assumptions, the probability is expressed as:
\[
P(p_D < p_0) =  E\left[ (1 - p_D)^{m-1}\right]
\]
where $p_D$ is the $p$-value associated with the data in the test statistic, and $p_0$ is the minimum of all $p$-values in the data going into null plots.
\end{lemma}

@@

%\red{pull the results for $P(p_D < p_{i,0})$ and $P(p_D < p_0)$ to the front, including the Figure on $1/m$ versus power. These results are independent of the regression setting - they don't make any assumptions in terms of regression.}

>>>>

\usepackage{setspace}
\doublespacing

@@

\begin{table}[hbtp]
%\renewcommand\thetable{}
\caption{Selection of 10 lineups for a person to evaluate in the simulation experiment with discrete covariate described in section 5.1 of the paper.} 
\centering
\begin{tabular}{c c c c  c c c}
\hline\hline
Difficulty& \multicolumn{3}{c}{parameter combination}& Number of evaluations &Total number  & number of lineups\\
 \cline{2-4}
level & $n$ & $\sigma$ & $\beta$ &required($n_{\gamma}$) & of lineups & randomly shown \\
\hline
easy&100& 5&8 & 1& 12 & 1\\
&100&12&16 &1&&\\
&300& 5&5 &1&&\\
&300&12&10 &1 &&\\
\hline
medium&100& 5&3 &203 & 9 &2\\
&300& 5&2, 3 & 97, 1&&\\
\hline
hard&100&12&3, 8, 10 & 277, 126, 23& 18 &6\\
&300& 5&1 & 371 &&\\
&300&12&3, 5& 375, 74 &&\\
\hline
mixed&100& 5&1, 5, 0& 214, 2, 73 & 21 &1\\
&100&12&1& 100& &\\
&300& 5&0 & 73&&\\
&300&12&7, 1& 2, 152&&\\
\hline
Total &&&&&60&10\\
\hline
\end{tabular}
\label{tbl:dist_lineup1}
\end{table}

\section{Proof of the Lemma}

@@

\section{Data Cleaning}

Following six approaches have been examined to clean the data.  Table \ref{tbl:summary} displays number of subjects and there total evaluations in each screening criteria. For our paper we picked criteria six.

\begin{enumerate}
\item {\bf Include all participants} and their evaluations.
\item Exclude all participants' evaluations, who did {\bf not} share their {\bf demographic information} (age, gender education level -- all three pieces of information are either missing or all present).
\item Exclude participants' records, if {\bf none of the evaluations}  correctly identified the data plot -- every participant was shown a range of `easy' lineups.
\item Include participants' records, if   {\bf at least 20 percent} of the evaluations are {\bf correct}  -- based on ten evaluations per participants, two correct evaluations are significant evidence against a person just guessing.
\item Include participants' records, if at least {\bf 50\% of all very easy (p-value $<$ 0.0002) lineups are correct} .
\item Use easy lineups as {\bf reference charts}: sample one easy lineup from a person's records. If that lineup is evaluated {\bf correctly}, include all (other) lineups of that person, otherwise exclude all lineup evaluations by this participant.
\end{enumerate}

The power curves obtained by fitting mixed model to the data for all the experiments applying different screening criteria are shown in Figure \ref{fig:power_screening_subject}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2.  % Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.70}{\includegraphics{power_screening.pdf}}
%       \caption{The overall power estimated from mixed model  for each screening criteria are shown. The corresponding power curve for conventional test is shown for comparison.}
%       \label{fig:power_screening}
%\end{figure}

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening_subject.pdf}}
       \caption{The subject specific power estimated from mixed model for all the six screening criteria are shown. The corresponding power curves for conventional test is shown for comparison.}
       \label{fig:power_screening_subject}
\end{figure}
